Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models

### 2020-06-11 analysis/paper/paper.Rmd
completed revisions through Results

The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River Basin, Pacific Northwest, USA. Within the interior Columbia River Basin two major runs of Chinook salmon (*Oncorhynchus tshawytscha*) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon enter freshwater earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in the freshwater before entering the ocean as smolts. Ocean-type Chinook salmon enter freshwater later (e.g. fall or winter) spawn lower in the watershed, and the juveniles may only spend between several weeks and six months in freshwater before migrating to the ocean. Here we focus on juvenile stream-type Chinook, in particular the summer rearing period during low flow. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River spring/summer-run is listed as threatened [@NOAA2016; @NOAA2016b]. Hereafter, we refer to both ESUs simply as Chinook salmon.

@@

1. Identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
1. Use paired fish and habitat measurements to elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density,
1. Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured (i.e., CHaMP sites within the Columbia River Basin),
1. Extrapolate capacity predictions at CHaMP sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
1. Validate estimates of carrying capacity from QRF across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).

@@

Fish and habitat data used in our study were collected from eleven watersheds within the interior Columbia River Basin, Pacific Northwest, USA (Figure \@ref(fig:wtsd-map)). The Columbia River Basin covers more than 668,000 km$^2$ draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia. The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) [@Volk2017] and were downloaded from https://www.champmonitoring.org. Data from the following `r as.character(english(n_distinct(hab_data$Watershed)))` CHaMP watersheds were used in this study: `r paste(sort(unique(hab_data$Watershed))[-length(unique(hab_data$Watershed))], collapse = ', ')` and `r sort(unique(hab_data$Watershed))[length(unique(hab_data$Watershed))]`. Juvenile density and abundance data were collected in a subset of `r as.character(english(n_distinct(fish_hab$Watershed)))` watersheds, at CHaMP survey reaches and were graciously provided by a number of projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].

@@

CHaMP sites are 200 m to 500 m reaches within wadeable streams across select basins within the interior Columbia River Basin selected based on a spatially balanced Generalized Random Tesselation Stratified sample selection algorithm [@Stevens1999; @Stevens2004]. Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, size (depth, width, discharge), substrate, temperature, and water quality.

@@

To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of Generalized Random Tesselation Stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. We fit two extrapolation models to each type of fish density, one that included watershed as covariate, for use in predicting capacity within CHaMP watersheds, and one that did not, for predicting everywhere else. We then made predictions of linear and areal capacity at all master sample sites throughout the interior Columbia River Basin, generally spaced about one kilometer apart. Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).

@@

Spawner-recruit data from several watersheds within the interior Columbia River Basin were compiled to validate the extrapolated QRF estimates of spring/summer Chinook salmon parr capacity. Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts (i.e. fall emigrants) from rotary screw traps. For the latter, estimates of parr were calculated using estimates of over-winter survival. A series of spawner-recruit functions were then fit to this data including Beverton-Holt, Ricker, and hockey stick, using the *FSA* package [@R-FSA] in R. Estimates of capacity from each of these spawner-recruit curves were compared with QRF estimates of capacity for the same regions.

@@

We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m). We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average august temperature, wetted width:depth ratio, percent fines less than 6mm, total percent fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \@ref(tab:select-mets-table)).

@@

A QRF model was fit using those metrics and the *quantregForest* package [@Meinshausen2006] in R [@Rsoftware2019] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity. After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \@ref(fig:rel-imp-figure)). Moreover, QRF models allow one to visually examine the marginal effect of each habitat covariate on the quantile of interest using partial dependence plots (PDP). These plots show the marginal effect of changing a single habitat covariate while maintaining all other covariates at their mean values (Figure \@ref(fig:pdp-figure)). However, given that many habitat metrics are somewhat correlated, these marginal effects are often not independent of one another, so care should be taken when interpretting them. After model fitting, the QRF model was used to predict habitat capacity at all CHaMP sites within the interior Columbia River basin.

@@

We fit a linear regression extrapolation using QRF-based predictions of capacity at all CHaMP sites as the response, and various GAAs as the independent variables. The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars). From this, we have estimates of capacity at every master sample point in the Columbia River Basin, which are located approximately one kilometer apart.

@@

<!-- While the Columbia River Basin is an excellent candidate for this type of extrapolation given the wealth of CHaMP data, application of this technique to areas with limited habitat data may be considerably more challenging. On-the-ground habitat data collection is costly, time-consuming, and often not reproducible; thereby limiting the effectiveness of this type of extrapolation.  Additionally, the extrapolation model relies on Global Available Attribute (GAA) data which ….  -->

@@

Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the Columbia River Basin, there is a pressing need to develop new tools for habitat analyses.  Unmanned Aerial Systems (UAS or drone, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency [@Jones2006; @Chabot2015]. UAS produce high-resolution, permanent data at a fraction of the cost of on-the-ground habitat sampling. Advances in imaging techniques (e.g., multispectral imaging) and post processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2018].

>>>>

The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River basin, Pacific Northwest, USA. Within the interior Columbia River basin two major runs of Chinook salmon (*Oncorhynchus tshawytscha*) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon enter freshwater earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in the freshwater before entering the ocean as smolts. Ocean-type Chinook salmon enter freshwater later (e.g. fall or winter) spawn lower in the watershed, and the juveniles may only spend between several weeks and six months in freshwater before migrating to the ocean. Here we focus on juvenile stream-type Chinook, in particular the summer rearing period during low flow. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River spring/summer-run is listed as threatened [@NOAA2016; @NOAA2016b]. Hereafter, we refer to both ESUs simply as Chinook salmon.

@@

1. Identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
1. Use paired fish and habitat measurements to elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density,
1. Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured (i.e., CHaMP sites within the Columbia River basin),
1. Extrapolate capacity predictions at CHaMP sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
1. Validate estimates of carrying capacity from QRF across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).

@@

Fish and habitat data used in our study were collected from eleven watersheds within the interior Columbia River basin, Pacific Northwest, USA (Figure \@ref(fig:wtsd-map)). The Columbia River basin covers more than 668,000 km$^2$ draining large portions of Idaho, Oregon, and Washington, and smaller portions of Montana, Nevada, Utah, and Wyoming, as well as the southeastern portion of British Columbia. The habitat data used to populate the QRF model were collected by the Columbia Habitat Monitoring Program (CHaMP) [@Volk2017] and were downloaded from https://www.champmonitoring.org. Data from the following `r as.character(english(n_distinct(hab_data$Watershed)))` CHaMP watersheds were used in this study: `r paste(sort(unique(hab_data$Watershed))[-length(unique(hab_data$Watershed))], collapse = ', ')` and `r sort(unique(hab_data$Watershed))[length(unique(hab_data$Watershed))]`. Juvenile density and abundance data were collected in a subset of `r as.character(english(n_distinct(fish_hab$Watershed)))` watersheds, at CHaMP survey reaches and were graciously provided by a number of projects, including the Integrated Status and Effectiveness Monitoring Project [@Volk2017].

@@

CHaMP sites are 200 m to 500 m reaches within wadeable streams across select basins within the interior Columbia River basin selected based on a spatially balanced Generalized Random Tesselation Stratified sample selection algorithm [@Stevens1999; @Stevens2004]. Habitat data within CHaMP sites were collected using the CHaMP protocol [@CHaMP2016] which calls for field data collection during the low-flow period, typically from June through October. CHaMP habitat data include, but are not limited to, measurements describing channel complexity, channel units, disturbance, fish cover, large woody debris, riparian cover, size (depth, width, discharge), substrate, temperature, and water quality.

@@

To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of Generalized Random Tesselation Stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. We fit two extrapolation models to each type of fish density, one that included watershed as covariate, for use in predicting capacity within CHaMP watersheds, and one that did not, for predicting everywhere else. We then made predictions of linear and areal capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart. Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).

@@

Spawner-recruit data from several watersheds within the interior Columbia River basin were compiled to validate the extrapolated QRF estimates of spring/summer Chinook salmon parr capacity. Some watersheds had direct estimates of parr, while some had estimates of smolts and pre-smolts (i.e. fall emigrants) from rotary screw traps. For the latter, estimates of parr were calculated using estimates of over-winter survival. A series of spawner-recruit functions were then fit to this data including Beverton-Holt, Ricker, and hockey stick, using the *FSA* package [@R-FSA] in R. Estimates of capacity from each of these spawner-recruit curves were compared with QRF estimates of capacity for the same regions.

@@

We categorized `r filter(hab_dict, MetricGroupName == "Visit Metric") %>% nrow()` habitat measurements collected using the CHaMP habitat protocol [@CHaMP2016] into `r n_distinct(hab_dict$MetricCategory[!is.na(hab_dict$MetricCategory)]) %>% english %>% as.character` habitat categories, and for each habitat covariate an MIC value was calculated based on the strength of association between the habitat covariate and the response variable, parr density (fish/m). We chose the following `r as.character(english(nrow(sel_hab_mets)))` CHaMP habitat covariates to fit the QRF model: wetted width, observed discharge, average August temperature, wetted width:depth ratio, percent fines less than 6 mm, total percent fish cover, channel unit frequency, standard deviation of the wetted depth, frequency of large wood in pools, percent riparian canopy cover, lower quantile of substrate size (D16) and braidedness (Table \@ref(tab:select-mets-table)).

@@

A QRF model was fit using those metrics and the *quantregForest* package [@Meinshausen2006] in R [@Rsoftware2019] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity. After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \@ref(fig:rel-imp-figure)). Moreover, QRF models allow one to visually examine the marginal effect of each habitat covariate on the quantile of interest using partial dependence plots. These plots show the marginal effect of changing a single habitat covariate while maintaining all other covariates at their mean values (Figure \@ref(fig:pdp-figure)). However, given that many habitat metrics are somewhat correlated, these marginal effects are often not independent of one another, so care should be taken when interpretting them. After model fitting, the QRF model was used to predict habitat capacity at all CHaMP sites within the interior Columbia River basin.

@@

We fit a linear regression extrapolation using QRF-based predictions of capacity at all CHaMP sites as the response, and various GAAs as the independent variables. The coefficients for the extrapolation model can be found in Table \@ref(tab:gaa-covars). From this, we have estimates of capacity at every master sample point in the Columbia River basin, which are located approximately one kilometer apart.

@@

<!-- While the Columbia River basin is an excellent candidate for this type of extrapolation given the wealth of CHaMP data, application of this technique to areas with limited habitat data may be considerably more challenging. On-the-ground habitat data collection is costly, time-consuming, and often not reproducible; thereby limiting the effectiveness of this type of extrapolation.  Additionally, the extrapolation model relies on Global Available Attribute (GAA) data which ….  -->

@@

Given the cost/extent of data necessary for QRF extrapolation in watersheds outside of the Columbia River basin, there is a pressing need to develop new tools for habitat analyses.  Unmanned Aerial Systems (UAS or drone, commonly) are gaining popularity in wildlife and ecosystem monitoring for their ease of use, safety, accessibility, and cost-efficiency [@Jones2006; @Chabot2015]. UAS produce high-resolution, permanent data at a fraction of the cost of on-the-ground habitat sampling. Advances in imaging techniques (e.g., multispectral imaging) and post processing (e.g., automation of data collection from imagery) are already demonstrating increase in the efficiency and accuracy of data collection [@Whitehead2014; @LeCun2015; @Weinstein2018].
